---
title: "Data Cleanup"
author: "Joy-El Talbot"
date: "2/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data pre-processing

Here, I load and create some summary statistics about the global.health data to let me reasonably reduce the dataset for the ALY6070 course.

```{r load_data}
setwd("~/Roux/2021Wb_ALY6070") # UPDATE for your working directory
covid19 <- read.csv("data/verified_globaldothealth_02-28-2021.csv", row.names=NULL)
```

### Summary statistics

A combination of summary() results, queries for unique values in each field, and counts of rows with a blank value were used to build a summary of the data contents for each column. These full results can be found in the modified data_dictionary file.
```{r summary_stats}
summary(covid19)

# example code for viewing unique values; this step was skipped if the summary reported all records as NAs
unique(covid19$events.outcome.value)

# example code to count number of rows with empty string values
sum(covid19$events.outcome.value == "")  # note that TRUE is coded as 1 and FALSE as 0
```

### Reduce table size by variable

Exclusion criteria:
1. All fields providing no distinguishing information (eg. all entries are NA)
2. Meta-data referencing other sources (eg. caseReference.additionalSources)
3. Meta-data with limited presence (eg. revisionMetadata.*)
4. Variable formatting fields with similar data elsewhere (eg. location.name)

```{r reduce_table_by_variable}
# fields to keep; see data_dictionary for details about all fields
keep = c("X_id", 
         "caseReference.sourceId",
         "caseReference.sourceUrl",
         "caseReference.verificationStatus",
         "demographics.ageRange.end",
         "demographics.ageRange.start",
         "demographics.ethnicity",
         "demographics.gender",
         "events.confirmed.date",
         "events.hospitalAdmission.date",
         "events.hospitalAdmission.value",
         "events.onsetSymptoms.date",
         "events.outcome.date",
         "events.outcome.value",
         "location.administrativeAreaLevel1",
         "location.administrativeAreaLevel2",
         "location.administrativeAreaLevel3",
         "location.country",
         "location.geoResolution",
         "location.geometry.latitude",
         "location.geometry.longitude",
         #"notes",
         "symptoms.status",
         "symptoms.values"
         )

covid19_new <- covid19[,keep]
# keeping notes separately to reduce file size
covid19_notes <- covid19[covid19$notes != "",c("X_id", "notes")] 

rm(covid19)

# export new data file
write.csv(covid19_new,
          file="data/reduced_globaldothealth_02-28-2021.csv",
          quote=TRUE,
          row.names=FALSE,
          col.names=TRUE)
write.csv(covid19_notes,
          file="data/notes_globaldothealth_02-28-2021.csv",
          quote=TRUE,
          row.names=FALSE,
          col.names=TRUE)
```